A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2025-02-13 manuscript/manuscript.qmd
COTA edits

**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J. Belli^6^, Sebastian Schneeweiss^1^, Shirley V. Wang^1^

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to the field of oncology. Studies in oncology come with their own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

@@

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The critical points considered in this step include an initial feasibility assessment of the data fitness[@rivera2022friends], including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

@@

The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results. If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies[@guidelinegeneral], and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: regulatory agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated @tbl-metrics.

@@

**Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).

@@

To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations. These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different. Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical], this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]

@@

Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability. The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.

@@

Given that a large number of factors could influence a difference between RCT results and emulated trial results, a broad range of sensitivity analyses will be conducted. This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients. All sensitivity will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.

@@

Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs. Due to intercurrent events, such as non-adherence or the crossover of a high proportion of patients from the control to the intervention arm, trial emulation are prone to be biased towards the null. Although this is also a common challenge in the analysis of RCTs[@rufibach2018], treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.

>>>>

**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J. Belli^6^, Ching-Kun Wang^6^, Sebastian Schneeweiss^1^, Shirley V. Wang^1^

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to the field of oncology. Studies in oncology come with their own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

@@

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The critical points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

@@

The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases). If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies,[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated @tbl-metrics.

@@

**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).

@@

To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations. These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different. Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]

@@

Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according to the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability. The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.

@@

Given that a large number of factors could influence a difference between RCT results and emulated trial results, a broad range of sensitivity analyses will be conducted. This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients. All sensitivity analyses will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.

@@

Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs. Due to intercurrent events, such as non-adherence or the crossover of a high proportion of patients from the control to the intervention arm, trial emulation are prone to be biased towards the null. Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
